SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2018-11-15 back.tex
reorg

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This is the template for submission to ISCA 2018
% The cls file is a modified from  'sig-alternate.cls'
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass{sig-alternate} 
\usepackage{mathptmx} % This is Times font

\newcommand{\ignore}[1]{}
\usepackage{fancyhdr}
\usepackage[normalem]{ulem}
\usepackage[hyphens]{url}
\usepackage{microtype}
\usepackage{fixltx2e}
\usepackage{array}
\usepackage{multirow}

% Always include hyperref last
\usepackage[bookmarks=true,breaklinks=true,letterpaper=true,colorlinks,linkcolor=black,citecolor=blue,urlcolor=black]{hyperref}

% Ensure letter paper
\pdfpagewidth=8.5in
\pdfpageheight=11in

%%%%%%%%%%%---SETME-----%%%%%%%%%%%%%
\newcommand{\iscasubmissionnumber}{NaN}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\fancypagestyle{firstpage}{
  \fancyhf{}
\renewcommand{\headrulewidth}{0pt}
  \fancyhead[C]{\normalsize{}} 
  \fancyfoot[C]{\thepage}
}

\pagenumbering{arabic}

%%%%%%%%%%%---SETME-----%%%%%%%%%%%%%
\title{SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher} 
\author{}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\maketitle
\thispagestyle{firstpage}
\pagestyle{plain}

\begin{abstract}
\textbf{Hardware prefetching has been introduced in modern processors as a way to hide cache latencies. 
An efficient prefetcher should be able to identify complex memory access patterns during a program execution. 
This enables it to fetch a block ahead of its demand access, potentially saving a cache miss.
}

\textbf{In this paper, we introduce perceptron learning to help make this prefetching decision efficiently. 
We have taken a state of the art prefetcher - Signature Path Prefetcher (SPP) as the underlying prefetch engine. 
The perceptron layer acts as a throttler to filter out the unnecessary prefetches recommended by SPP. 
We have also explored a range of features that can be used to train the perceptron layer. 
Our results show that SPP-Perc improves performance on the memory intensive subset of the SPEC 2017 Benchmark suite by 6.84\% on single-core and by 11.9\% on multi-core traces, as compared to just SPP.
We also demonstrate that the performance gained from using our efficient throttler scales better with increasing number of cores.
}
\end{abstract}

\section{Introduction}
Since past few decades, processor and memory scaling has been enabled with different philosophies in mind. 
While processor scaling focused on speed improvements, memory scaling was centred around increasing the storage capacity. 
The process technology reaching scaling limitations led to Memory Wall\cite{MemWall} - the increasing performance gap between the processor speed and the memory speeds. 
Various techniques have been developed to combat this widening gap. 
Data prefetching is one such important technique. 
Data prefetching exploits the fact that in most applications, memory accesses have a well-defined pattern.

An ideal prefetching scheme involves capturing this memory access patterns to predict future prefetches in a timely manner. 
Other important decisions faced by the prefetcher are when to bring in the prefetch block of data, which level of cache hierarchy to place the block in and finally, what block to evict to accommodate the incoming block. 
Memory accesses patterns can be as simple as fetching the next cache line or a complex but predictable pattern, like in pointer chasing applications. 
Predicting this often requires a trade-off between coverage and accuracy. 
For a prefetcher to capture highly irregular access patterns, it needs to be highly aggressive (high coverage). 
That also means the prefetcher would end up recommending a lot of useless prefetches, leading to cache pollution (low accuracy). 
This effect would get even more pronounced in multi-core scenario where the last level cache is a shared resource. 
Conversely, a conservative prefetcher would be highly accurate but might not make enough useful predictions.

To this effect we propose SPP-Perc <come up with a catchy name>. 
It is an enhancement to the existing state of the art prefetcher and overcomes the coverage vs accuracy trade-off in a robust manner. 
The idea is to modify SPP design to make it as aggressive as it can get. 
This enables SPP to capture complex memory access patterns hence leading to increased coverage.
This comes at a cost of increased DRAM traffic, hence reduction in overall IPC.
The prefetch suggestions recommended by SPP pass through the perceptron based throttler. 
Over time, perceptron layer learns to correlate the prefetches recommended by SPP with various available features, leading to elimination of useless prefetches. 
This leads to an overall increased accuracy of the prefetch scheme and reduction in DRAM traffic.

\vspace{1ex}This paper talks about SPP-Perc, its merits, analysis and future scope for expansion. 
The contributions made by this paper are:

\begin{itemize}
\item This is the first time where a truly on-line learning based neural network model has been used when it comes to hardware data prefetching. 
The prior works in this area either relied on program semantics\cite{Semantics} or were application specific\cite{Datacenter} or could not yield a significant improvement over the baseline prefetcher used\cite{BadPerc}.

\item Implementing the pereptron throttler in a robust and accurate practical prefetching mechanism. 
The overall scheme yields increased accuracy as compared to the other state of the art prefetchers.
Moreover, the perceptron throttler learns to adapt itself to shared resource constraints - leading to an even increased performance in multi-core and lower capacity / bandwidth environments.

\item Examining the perceptron weights doing a statistical study to quantify the contribution made by each feature.
We have also used this idea to develop a comprehensive set of features where each feature contributes some `new' information to the perceptron.
The details are explained under methodology, in Section XX.

\item Cross-validating the design across benchmarks. 
All the development and parameter tuning was done on the SPEC 2017 benchmark. 
The results have been shown to be consistent across unseen benchmarks.
More details can be found in the results section (Section XX).

%\item Designing the prefetcher keeping SPEC 2017 benchmark suite in mind. 
%We have used SPEC 2017 to compare SPP-Perc against the highest performing prefetchers. 
%We have also used our prefetcher to characterize memory intensive benchmarks of the SPEC 2017 suite, with respect to prefetching.

%\item We came up with an efficient model to use the perceptron output as a representation of the prefetch confidence using an on-line histogram technique. 
%We show how the simple sum can replace the existing mechanism which was essentially based on counting the number of hits for a particular entry in SPP.

\end{itemize}

We tested SPP-Perc on Champsim, an open source simulator which models an Out-of-Order CPU using. 
For evaluation, we used SPEC 2017 benchmarks. 
On a single core configuration, SPP-Perc increases performance of the system by 39.5\% as compared to the baseline with no prefetching, on the memory intensive subset of the benchmark.
This outperforms the next best prefetcher, SPP,  by 6.84\%. 
On a multi-core system running a mix of memory intensive subset of the SPEC 2017 traces, SPP-Perc saw an improvement of 11.9\% over SPP for 4-core system and 10.67\% for 8-core system.

In the results part in section XX, we show a detailed analysis of the speedup achieved.
We also used SPEC 2006 and CloudSuite benchmarks for cross-validation of the prefetcher implementation.

The rest of the paper is organized as follows.
Section 2 talks about background work and recaptures the SPP architecture
Section 3 discusses the detailed implementation of the perceptron mechanism and the features used.
Section 4 consists of the evaluation methodology and explores the feature space for perceptron learning.
Section 5 is the performance evaluation of our prefetcher and Section 6 concludes the paper.

@@

The idea of prefetching has been around since Instruction Stream Buffers were proposed by Jouppi.
\cite{ISB}
Some of the earliest data prefetchers developed to try to identify memory access patterns with a constant stride pattern.
\cite{Smith} 
\cite{Baer}
\cite{Stride}
Major research was dedicated to stride distance and depth prediction.
\cite{Decoupled}
\cite{Adaptive}
Newer prefetchers aimed at correlating predictions with the past memory access addresses.
\cite{Address_Correlated} 
\cite{AMPM}

With time newer and more intricate prefetchers were introduced.
\cite{Wenisch_Temporal_Streaming}
\cite{Stealth}
\cite{Feedback_Directed}
\cite{Coordinated}
\cite{Bandwidth_Efficient}
\cite{Pacman}
\cite{TLB}
\cite{Linearizing}
\cite{Sandbox}
\cite{VLDP}
\cite{DoL}
\cite{Domino}
A whole class of streaming prefetchers developed on the lines of Spatial / Temporal Memory Streaming.
\cite{Spatial_Pattern}
\cite{SMS}
\cite{Temporal_Instruction_Fetch}
\cite{Off_Chip}
\cite{STMS}
\cite{SMS_JILP}

@@

A related line of work also explores cache replacement - insertion policies and dead block prediction, with an eye towards prefetching. 
\cite{DB_Pred}
\cite{Cache_Burst}
\cite{KPC}
\cite{Harmony}
Nori \textit{et.al.} introduced prefetching with  respect to runtime criticality.
\cite{CATCH}
\newline
\newline
\textit{Machine Learning and Prefetching:}
Peled \textit{et.al.} introduce interesting ideas for on-line Reinforcement Learning and dynamically scaling the magnitude of feedback given to the baseline prefetcher.
\cite{Semantics}
The biggest challenge here is that it relies on compiler support for getting features to build the context.

@@

%% NOTE: The below paper is literally the same idea as ours. But poor implementation and no good results. How much to bash this paper?
Finally, the paper on Data Cache Prefetching with Perceptron Learning\cite{BadPerc} talks about the idea of two step prefetching.
The first step is an existing baseline prefetcher while the second step is a perceptron based throttler.
The paper has lots of limitations in terms of design and implementation.
That reflects in the fact that it did not lead to any significant performance gain over even the basic prefetchers like the Stride Prefetcher\cite{Stride} and the Markov Prefetcher\cite{Markov}.

@@

The work was further extended by Multiperspective Reuse Prediction\cite{Multiperspective}, which introduced a varied set of parametric features. 
Although design space exploration for feature and parameter selection was a non-trivial task, once fixed, it could give a cache replacement predictor highly suitable for the given set of applications.
\newline
\newline
\textit{Deviations From Actual Perceptrons:} Traditionally, a perceptron prediction involves multiplying the vector of input features: F\textsubscript{1xN}, with the corresponding weight vector: W\textsubscript{Nx1} in a dot product fashion to obtain the sum y\textsubscript{out}. 
Here what we use is a perceptron-like structure. 
The feature is used to hash into the weight of perceptrons and the retrieved weights are added straight away. 
This can be seen a dot product of $\vec{1}$\textsubscript{1xN} with W\textsubscript{Nx1}. 
This way, the perceptron algorithm doesn't introduce any multiplication operations in the inference or training process. 
Hence what we adapt in this work, is a perceptron-like learning algorithm as it involves the same principle involved in perceptron inferencing and training. 
For simplicity, we will call our implementation as perceptron implementation.
\newline
\newline
\textit{Perceptron Update Rule:} In all the above implementations, a uniform perceptron update principle is being followed. 
The weights need to be updated if the prediction was wrong or the predicted sum does not cross a certain threshold. 
After a certain training period, these weights are proportional to the probability of the outcome being true. 
This training threshold makes sure that the perceptrons are trained till a certain level of confidence and yet they are not over-trained to the given set. 
Care is taken that the weights saturate at certain positive and negative values so that they remain confined to the bit-width. 
The same rules will be applied in the discussion done in section XX on SPP-Perc learning algorithm.

@@

\subsection{Signature Path Prefetcher Architecture}
Since the work we have done builds heavily on SPP\cite{SPP}, we are presenting a brief recap of the ideas introduced in the original paper, especially with an eye towards their relevance in our work. 
Signature Path Prefetcher (SPP) is a lookahead prefetcher and consists of following structural components:
\newline
\newline
\textbf{Signature Table:} ST keeps track of 256 most recently accessed pages. 
It is meant to capture memory access patterns within a page boundary. 
SPP indexes into an entry of ST using the page number. 
For each entry corresponding to a page, ST stores the `last block offset' and `old signature'. 
Last block offset is the block offset of the last memory access of that given page. 
The block offset is calculated with respect to the page boundary. 
The signature is a 12-bit compressed representation of the past few memory accesses for that page. 
The signature is calculated as:
$$New Signature = (\,Old Signature << 3 bits\,) \;\;XOR\;\; (\,Delta\,)$$ 
In case a matching page entry is found, the stored signature in the corresponding entry is retrieved.
\newline
\newline
Consider the case that incoming page number 10 with a block offset 3 finds a match in the ST. 
The retrieved pattern signature is 0x30 and the Last Offset is 1. 
Since now there is the Last Block Offset (1), Incoming Block Offset (3), Old Signature (0x30), and New Signature calculated as per above equation (0x182), SPP can infer non-speculatively that the given pattern of memory accesses (as captured in the Old Signature) leads to the particular Delta. 
In general, Delta is defined as the difference between the prefetch suggestion block and the initial block which triggered the prefetch. 
In this case, since SPP is in learning phase, it is defined as the difference between the Incoming Block Offset and the Last Block Offset (+2 in this case). 
This, in turn generates the new memory pattern (New Signature). 
This newly learned signature pattern and the delta is stored in the Pattern Table.
\newline
\newline
\textbf{Pattern Table:} The PT is indexed by the signature generated from ST. 
PT holds predicted delta patterns and their confidence estimates. 
Each entry indexed by the signature holds up to 4 unique delta predictions. 
This is implemented by making PT as a 4-way associative table. 
To measure an estimate of a delta confidence, PT implements two counters: C\textsubscript{sig} for each entry indexed by the signature and C\textsubscript{delta} for each delta entry for a given signature. 
More details of this on-line score-keeping mechanism is described under Confidence Tracking discussion.

@@

In the lookahead mode, SPP looks at the last generated delta(s), chooses the one with the highest confidence and speculatively generates the new effective address and the new signature. 
Using these new parameters, it re-accesses the PT to generate further prefetches. 
SPP keeps on repeating this cycle of accessing PT and updating the signature based on highest confidence prefetch from the last iteration. 
These iteration count till which SPP manages to predict prefetch entries in the lookahead manner is characterized as its `depth'. 
While doing this, SPP also keeps compounding the confidence in each depth. 
Thus as depth increases, overall confidence keeps decreasing. 
This makes sense since we are going deeper into the speculation path. 
When the confidence falls below a pre-defined threshold, SPP stops further lookahead iterations
\newline
\newline
\textbf{Global History Register:} One of the innovative ideas introduced in SPP was the Global History Register, which bootstraps learning across page boundaries.
GHR is a small 8-entry fully associative structure. 
Whenever a prefetch suggestion from the Pattern Table crosses page boundaries, instead of completely discarding it, GHR saves that information for inter-page learning.
Since the SPP-Perc implementation in this paper doesn't affect GHR functionality in any ways, we won't be going into more details about GHR.
\newline
\newline
\textbf{Prefetch Filter:} SPP also introduced the concept of Prefetch Filter - a 1024 entry 1-way associative table which keeps a record of last few entries prefetched by SPP. 
This proves useful for following reasons:
\begin{itemize}
    \item \textbf{Eliminating duplicate prefetches}\newline 
    Any subsequent prefetches recommended by SPP mechanism is checked against the existing entries in Prefetch Filter. 
If a valid entry is found, it means that the particular cache line is already either in L2 Cache or being fetched. 
Hence there is no need to issue a duplicate prefetch.

@@

Here d is the lookahead depth. 
For d = 1, when SPP is in non-speculative mode, P\textsubscript{0} can be thought of as 1. 
Once SPP enters into speculative mode, it calculates the path confidence as a product of confidence of the individual prefetches encountered along that path, further scaled using $\alpha$. 
The final P\textsubscript{d} is thresholded against prefetching-threshold (T\textsubscript{p}) to decide whether to prefetch or not. 
If yes, then P\textsubscript{d} is thresholded against a numerically bigger fill-threshold (T\textsubscript{f}) to decide whether the prefetch should be sent to L2 Cache (high confidence prefetch) or Last Level Cache (low confidence prefetch)
\newline<Insert a comprehensive figure about SPP structures>
\newline<Insert another detailed picture about SPP data-path flow>

@@

\section{SPP-Perc}

\subsection{Changes made to original SPP}
To modify the SPP Design to suit our scheme, the following changes were made:

\begin{itemize}
\item \textbf{Original Thresholds discarded}\newline
In the original SPP design, an internal confidence mechanism was developed which was compared against a preset PF\_THRESHOLD (T\textsubscript{p}) to decide whether to accept or reject the prefetch suggestion.
Furthermore, the accepted suggestions were thresholded against FILL \_THRESHOLD (T\textsubscript{f}) to decide the fill level: L2 Cache vs Last Level Cache.
T\textsubscript{p} and T\textsubscript{f} were set to 25 and 90 respectively, on the scale of 0 to 100.
In SPP-Perc, the perceptron sum is used to make these decisions. 
Hence these two thresholds are no longer needed

\item \textbf{Looking at L2 MSHR Queue while prefetching}\newline
Original SPP would not consider the slots available in the L2 MSHR queue. 
It had an assumption that at no time can the number of prefetches suggested by SPP could exceed the capacity of the queue. 
While it was not proven in the paper, the internal confidence mechanism along with the thresholds made sure that the above assumption was maintained. 
In SPP-Perc since there can be no assumption on the degree of aggressiveness - because SPP is being tuned to its lowest thresholds, we explicitly check that at no point should the number of suggested prefetches exceed L2 MSHR queue.

\item \textbf{Enhanced Filter}\newline
Original SPP introduces a concept of Prefetch Filter, which keeps a track of up to 1024 L2 Cache prefetches suggested by SPP. 
The Filter was modified to store meta-data like program counter, memory access address etc. 
This is needed to train the perceptron when the result of that particular prefetch gets available. 
At training time, this data is used to retrieve the state of the prefetcher when the prediction was made.

\item \textbf{Reject Filter introduced}\newline
The Prefetch Filter mentioned above keeps a track of the prefetches that actually happened in the L2 Cache. 
Correspondingly, we introduced a reject filter. 
It keeps a track of the prefetches that were suggested by the base SPP engine but were rejected by the perceptron. 
In case it is detected that a future demand fetch could have used this rejected prefetch, the perceptron weights are updated to reflect the same

\end{itemize}

\subsection{The Perceptron Layer}
As discussed earlier, the perceptron layer was introduced to act as a throttler to the underlying SPP engine. 
This section explores the details of the perceptron layer and the range of features that are ultimately used by the perceptron for correlating the prefetches.

The neural network weights are stored in separate tables for each feature.
The indexing of each feature is done using different number of bits.
At most 12 bits of a feature are used to index into the perceptron table.
Certain features require more resolution power \textit{i.e.}, full 12 bits of indexing.
On the other hand, some features requiring lower resolution required as low as 7 bits of indexing.
The variable indexing was determined by studying the features and fine-tuned empirically so as to achieve a good accuracy vs area trade-off.
Exact details are mentioned under `Area Overheads' in Section XX.

A single entry in the table corresponds to a perceptron weight.
Each weight is a 5 bit counter - saturating at -16 and +15. 
Each feature gets its own dedicated table. 
Our proposed design uses 9 features. 
As the program execution begins, all the weights are initialized to 0.
\newline \newline
\textbf{INFERENCING}\newline
In the Champsim simulator, the prefetching method is triggered on every L2 Cache demand access. 
At that point, the prefetcher has the option to do or not do prefetch - in case it does, then how many cache lines to prefetch. 
The suggested prefetches can be places either in L2 Cache or in Last Level Cache. 
The prefetcher gets to decide this, and is usually done based on an internal confidence mechanism.

When SPP gets triggered, it suggests up to 4 (=Associativity of the Pattern Table) candidates for prefetch based on the demand access address. 
In each subsequent lookahead depth, SPP suggests up to 4 more prefetch candidates, as explained in section\_\_. 
All these recommended prefetches are tested in the perceptron throttler before they can access the Prefetch Filter, as in the original SPP data-path.

The perceptron looks at the micro-architectural state, or `features', at that instant. 
Each feature indexes into the hash of up to 4096-entry table dedicated for that feature. 
Feature \#1 indexes into a particular entry in table \#1 and so on.

Once all the weights are retrieved, they are added together, much like the traditional perceptron dot product. 
On obtaining the overall sum, it is thresholded based on the preset threshold PERC\_THRESHOLD\_LO. 
Only the prefetch candidates with the perceptron sum higher that the threshold qualify for prefetching. 
The prefetches that qualify the perceptron throttler stage go through the Prefetch Filter stage, just as in the SPP design.
The Prefetch Filter eliminates any redundant prefetch suggestions.
In SPP-Perc, all the meta-data that is required to recreate the features is also stored in the Prefetch Filter. 
At a later stage, when the feedback of the current prefetched line is available, the stored data can be used to train the perceptron.

The above generated perceptron sum is also used to decide L2 Cache vs Last Level Cache placement of the prefetch cache line. 
All prefetch candidates qualified till this stage are thresholded against PERC\_THRESHOLD\_HI to decide the fill level. The two thresholds: PERC\_THRESHOLD\_LO and PERC \_THRESHOLD\_HI are empirically set.

The perceptron sum can also be seen as the sum of the individual contribution per feature.
The value of each contribution corresponds to the extent of confidence of the final output with that given feature.
By summing the individual contributions, the final perceptron sum denotes the overall confidence for that prefetch suggestion. 
In a way, by thresholding the perceptron sum on two different thresholds, we are diving the confidence scale of the prefetch suggestion into three bins.
The first bin corresponding to the lowest confidence leads to prefetch candidate being rejected.
The next bin corresponds to prefetches with a moderate confidence level. 
Such prefetches are directed to fill the bigger Last Level Cache and potentially not pollute the more scarce L2 Cache.
The highest confidence prefetches are filled in the L2 Cache.

In addition to the Prefetch Filter mentioned above, SPP-Perc also maintains a `Reject Filter'. 
Reject filter is a 1024-entry deep 1-way associative table. 
In case that a prefetch suggestion is rejected by perceptron throttler, it instead gets logged into the Reject Filter. 
The Reject Filter keeps track of 1024 entries that were suggested by the Pattern Table but got rejected by the perceptron throttler. 
This can be used to train the perceptron to avoid false negatives \textit{i.e.}, cases where prediction was to reject the prefetch but the prefetch would have been useful.

As previously stated, an enhancement to the original SPP design includes storing all the meta-data describing the program state at the instant of prefetching, in the Prefetch Filter. 
This information is useful when the perceptron needs to be updated subsequently. 
\newline \newline
\textbf{TRAINING}\newline
In the prefetching environment, feedback of a prefetch is received whenever there is an eviction or a demand access from L2 Cache. 
This acts as a trigger to update all the baseline SPP entries and to also train the perceptron throttler. 
Training of the perceptron throttler involves restoring the meta-data that was stored in Prefetch Filter or the Reject Filter. 
This is done by indexing into the corresponding filters using the cache line address of the block being used to train. (10-bits indexing + 6-bits tag matching). 
Once the state of the program at the time of prefetching is available, it is used to index into the perceptron weights table.

If a demand access block which triggers the training  was tagged as a valid prefetch in the Prefetch Filter, it means that the earlier prefetch prediction was correct. 
In that case the perceptron weights are incremented by 1 if the predicted sum does not cross a pre-defined threshold. 
 In case it is a cache block eviction that led to training and the corresponding valid entry was found in the Prefetch Filter, it shows that the prediction made by the perceptron was wrong. 
Perceptron should have ideally rejected the prefetch suggestion as a low-confidence prefetch. 
Here the weights are decremented by 1 to reflect the same. 
In either case, weights are saturated at -16 or +15.

A secondary training mechanism also kicks in during demand fetches. 
Before the demand access triggers the next set of prefetches, it checks the Reject Filter for the valid entry. 
In case it is a hit, that means that the corresponding cache line was initially suggested by the underlying SPP engine but rejected by the perceptron throttler. 
This means that perceptron should have been more confident about that particular prefetch. 
Once such a scenario is identified, the state of the execution at the time of prefetch is retrieved from the Reject Table. 
The retrieved data is used to index into the various weights tables of the perceptron and the corresponding values are updated by +1, saturating between and -16 and +15. 
This update reflects increased confidence for the prediction corresponding to that prefetch.

This mechanism allows us to solve the classical problem of exploiting `lost opportunity'. 
In prior perceptron based implementations (and in general, prefetching algorithms), there is usually no way of knowing the result of not prefetch a particular line.
Our two-step SPP-Perc allows us to overcome that issue in a simple manner.

\subsection{Features used by Perceptron}

Here we discuss the various features which help correlate the prefetching decision with the program behaviour
\begin{itemize}
\item \textbf{Base address} of the demand access which triggered the prefetch. 
Since prefetch are triggered from the L2 demand access, they tend to be correlated with the triggering address.

\item \textbf{Cache line} and \textbf{Page address}: These two separate features are derived from the base address which triggered the prefetch, as follows: base\_addr >> LOG2\_BLOCK \_SIZE and base\_addr >> LOG2\_PAGE\_ SIZE respectively.
The idea behind using three different shifted versions of the same feature is that it allows us to look into a wider range of bits than with a single version.
It also helps give more importance to the overlapping bits and lesser importance to most and least significance bits.
This approach can also eliminate destructive interference that can be caused by directly folding the bits into half.

\item \textbf{Confidence}: The integer confidence on the scale of 0 to 100 which was used in the original SPP design.
While the original confidence is not used directly for decision making, it can still contribute to the final decision made.

\item \textbf{Program Counter XOR Depth}: PC is for the instruction which triggered the prefetch chain.
Depth refers to the iteration count of the lookahead stages.
In general, PC is not considered as a good basis for doing lookahead prefetching as all the prefetches with depth >= 1 gets aliased onto the same PC which won't be true in actual demand fetch case.
This feature resolves a PC into a different value for each lookahead depth of SPP speculation, giving a more accurate correlation in lookahead cases.
This is akin to the concept of Virtual Program Counters\cite{VPC} introduced by Kim \textit{et. al.} for indirect branch prediction.

\item \textbf{PC\textsubscript{1} XOR PC\textsubscript{2}>>1 XOR PC\textsubscript{3}>>2}: Here $PC_i$ refers to the last $i^{th}$ PC before the instruction that triggered the current prefetch. 
Hashing together the last three PCs tell SPP-Perc about the path that led to the current demand access and helps capture and branching information of the current basic block.
PCs are shifted in the increasing order of history before being hashed together.
This is done to avoid the resultant value of zero in case 2 or more PCs are the same. 
Additionally, obfuscating the information as it gets old allows us to get a wider and an approximate look into the program history.

\item \textbf{Program Counter XOR Delta}: This tells us if a given PC favours particular value(s) of delta.
As noted earlier, while PC in itself does not much useful information, this hash resolves PC into different values based on the tendency of that PC to favour a certain delta.
This way, dynamic nature of different instances of the same memory access instruction can be captured here.

\item \textbf{Current Signature XOR Delta}: It can be recalled from the discussion of SPP in section XX that the new signature is generated using the old signature and the blobk delta.
While `Current Signature XOR Delta' is not the exact formula for generating the future signature, it gives an approximate idea of the path that the combination of these two values can lead to.

%%%%%%% Below are the features rejected when downsizing 14 -> 9
%\item \textbf{Current Signature}: The 12-bit signature used by SPP to index into the Pattern Table
%\item \textbf{Delta}: The signed version of the difference in the block address that triggers the prefetch and the block address of the data being prefetched
%\item \textbf{PC}: Program Counter of the Load instruction that triggers that particular prefetch
%\item \textbf{Base address XOR Delta}: If any particular value(s) of delta are highly favoured by a given address which triggered the prefetch, this composite feature can capture that information
%\item \textbf{Cache line XOR Delta}: This feature captures if any particular delta is highly favoured by a given cache line address
%\item \textbf{Confidence XOR Delta}: This gives a combined insight into that fact that certain delta-confidence pairs predicted by SPP might correlate more with the prefetch outcome
%%%%%%%%%%%%%%%%%%%%%%

\end{itemize}

>>>>

@@

The idea of prefetching has been around since Instruction Stream
Buffers were proposed by Jouppi.  \cite{ISB} Some of the earliest data
prefetchers developed to try to identify memory access patterns with a
constant stride pattern.  \cite{Smith,Baer,Stride} Major research was
dedicated to stride distance and depth prediction.
\cite{Decoupled,Adaptive} Newer prefetchers aimed at correlating
predictions with the past memory access addresses.
\cite{Address_Correlated,AMPM}

With time newer and more intricate prefetchers were
introduced~\cite{Wenisch_Temporal_Streaming,Stealth,Feedback_Directed,Coordinated,Bandwidth_Efficient,Pacman,TLB,Linearizing,Sandbox,VLDP,DoL,Domino}.
A whole class of streaming prefetchers developed on the lines of
Spatial / Temporal Memory
Streaming~\cite{Spatial_Pattern,SMS,Temporal_Instruction_Fetch,Off_Chip,STMS,SMS_JILP}.

@@

A related line of work also explores cache replacement - insertion
policies and dead block prediction, with an eye towards
prefetching~\cite{DB_Pred,Cache_Burst,KPC,Harmony}.  Nori
\textit{et.al.} introduced prefetching with respect to runtime
criticality~\cite{CATCH}.

\textit{Machine Learning and Prefetching:} Peled \textit{et.al.}
introduce interesting ideas for on-line Reinforcement Learning and
dynamically scaling the magnitude of feedback given to the baseline
prefetcher.  \cite{Semantics} The biggest challenge here is that it
relies on compiler support for getting features to build the context.

@@

%% NOTE: The below paper is literally the same idea as ours. But poor implementation and no good results. How much to bash this paper?
Finally, the paper on Data Cache Prefetching with Perceptron
Learning~\cite{BadPerc} talks about the idea of two step prefetching.
The first step is an existing baseline prefetcher while the second
step is a perceptron based throttler.  The paper has lots of
limitations in terms of design and implementation.  That reflects in
the fact that it did not lead to any significant performance gain over
even the basic prefetchers like the Stride Prefetcher\cite{Stride} and
the Markov Prefetcher\cite{Markov}.

@@

The work was further extended by Multiperspective Reuse
Prediction\cite{Multiperspective}, which introduced a varied set of
parametric features.  Although design space exploration for feature
and parameter selection was a non-trivial task, once fixed, it could
give a cache replacement predictor highly suitable for the given set
of applications.

\textit{Deviations From Actual Perceptrons:} Traditionally, a
perceptron prediction involves multiplying the vector of input
features: F\textsubscript{1xN}, with the corresponding weight vector:
W\textsubscript{Nx1} in a dot product fashion to obtain the sum
y\textsubscript{out}.  Here what we use is a perceptron-like
structure.  The feature is used to hash into the weight of perceptrons
and the retrieved weights are added straight away.  This can be seen a
dot product of $\vec{1}$\textsubscript{1xN} with W\textsubscript{Nx1}.
This way, the perceptron algorithm doesn't introduce any
multiplication operations in the inference or training process.  Hence
what we adapt in this work, is a perceptron-like learning algorithm as
it involves the same principle involved in perceptron inferencing and
training.  For simplicity, we will call our implementation as
perceptron implementation.

\textit{Perceptron Update Rule:} In all the above implementations, a
uniform perceptron update principle is being followed.  The weights
need to be updated if the prediction was wrong or the predicted sum
does not cross a certain threshold.  After a certain training period,
these weights are proportional to the probability of the outcome being
true.  This training threshold makes sure that the perceptrons are
trained till a certain level of confidence and yet they are not
over-trained to the given set.  Care is taken that the weights
saturate at certain positive and negative values so that they remain
confined to the bit-width.  The same rules will be applied in the
discussion done in section XX on SPP-Perc learning algorithm.

@@

\subsection{Signature Path Prefetcher Architecture}
Since the work we have done builds heavily on SPP\cite{SPP}, we are
presenting a brief recap of the ideas introduced in the original
paper, especially with an eye towards their relevance in our work.
Signature Path Prefetcher (SPP) is a lookahead prefetcher and consists
of following structural components:

\textbf{Signature Table:} ST keeps track of 256 most recently accessed
pages.  It is meant to capture memory access patterns within a page
boundary.  SPP indexes into an entry of ST using the page number.  For
each entry corresponding to a page, ST stores the `last block offset'
and `old signature'.  Last block offset is the block offset of the
last memory access of that given page.  The block offset is calculated
with respect to the page boundary.  The signature is a 12-bit
compressed representation of the past few memory accesses for that
page.  The signature is calculated as:
$$New Signature = (\,Old Signature << 3 bits\,) \;\;XOR\;\; (\,Delta\,)$$ 
In case a matching page entry is found, the stored signature in the corresponding entry is retrieved.

Consider the case that incoming page number 10 with a block offset 3
finds a match in the ST.  The retrieved pattern signature is 0x30 and
the Last Offset is 1.  Since now there is the Last Block Offset (1),
Incoming Block Offset (3), Old Signature (0x30), and New Signature
calculated as per above equation (0x182), SPP can infer
non-speculatively that the given pattern of memory accesses (as
captured in the Old Signature) leads to the particular Delta.  In
general, Delta is defined as the difference between the prefetch
suggestion block and the initial block which triggered the prefetch.
In this case, since SPP is in learning phase, it is defined as the
difference between the Incoming Block Offset and the Last Block Offset
(+2 in this case).  This, in turn generates the new memory pattern
(New Signature).  This newly learned signature pattern and the delta
is stored in the Pattern Table.

\textbf{Pattern Table:} The PT is indexed by the signature generated
from ST.  PT holds predicted delta patterns and their confidence
estimates.  Each entry indexed by the signature holds up to 4 unique
delta predictions.  This is implemented by making PT as a 4-way
associative table.  To measure an estimate of a delta confidence, PT
implements two counters: C\textsubscript{sig} for each entry indexed
by the signature and C\textsubscript{delta} for each delta entry for a
given signature.  More details of this on-line score-keeping mechanism
is described under Confidence Tracking discussion.

@@

In the lookahead mode, SPP looks at the last generated delta(s),
chooses the one with the highest confidence and speculatively
generates the new effective address and the new signature.  Using
these new parameters, it re-accesses the PT to generate further
prefetches.  SPP keeps on repeating this cycle of accessing PT and
updating the signature based on highest confidence prefetch from the
last iteration.  These iteration count till which SPP manages to
predict prefetch entries in the lookahead manner is characterized as
its `depth'.  While doing this, SPP also keeps compounding the
confidence in each depth.  Thus as depth increases, overall confidence
keeps decreasing.  This makes sense since we are going deeper into the
speculation path.  When the confidence falls below a pre-defined
threshold, SPP stops further lookahead iterations.

\textbf{Global History Register:} One of the innovative ideas
introduced in SPP was the Global History Register, which bootstraps
learning across page boundaries.  GHR is a small 8-entry fully
associative structure.  Whenever a prefetch suggestion from the
Pattern Table crosses page boundaries, instead of completely
discarding it, GHR saves that information for inter-page learning.
Since the SPP-Perc implementation in this paper doesn't affect GHR
functionality in any ways, we won't be going into more details about
GHR.

\textbf{Prefetch Filter:} SPP also introduced the concept of Prefetch
Filter - a 1024 entry 1-way associative table which keeps a record of
last few entries prefetched by SPP.  This proves useful for following
reasons:
\begin{itemize}
\item \textbf{Eliminating duplicate prefetches}\newline Any subsequent
  prefetches recommended by SPP mechanism is checked against the
  existing entries in Prefetch Filter.  If a valid entry is found, it
  means that the particular cache line is already either in L2 Cache
  or being fetched.  Hence there is no need to issue a duplicate
  prefetch.

@@

Here d is the lookahead depth.  For d = 1, when SPP is in
non-speculative mode, P\textsubscript{0} can be thought of as 1.  Once
SPP enters into speculative mode, it calculates the path confidence as
a product of confidence of the individual prefetches encountered along
that path, further scaled using $\alpha$.  The final
P\textsubscript{d} is thresholded against prefetching-threshold
(T\textsubscript{p}) to decide whether to prefetch or not.  If yes,
then P\textsubscript{d} is thresholded against a numerically bigger
fill-threshold (T\textsubscript{f}) to decide whether the prefetch
should be sent to L2 Cache (high confidence prefetch) or Last Level
Cache (low confidence prefetch)

<Insert a comprehensive figure about SPP structures>

<Insert another detailed picture about SPP data-path flow>

@@
